---
title: "R basics"
date: "Last updated: `r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

## First steps

This tutorial assumes that you have already installed R and RStudio.
Some good resources to get a basic familiarity with the setup of R and some key definitions are [A (very) short introduction to R](https://github.com/ClaudiaBrauer/A-very-short-introduction-to-R/blob/master/documents/A\%20(very)\%20short\%20introduction\%20to\%20R.pdf) and sections 1-4 of [How to make any plot in ggplot2?](http://r-statistics.co/ggplot2-Tutorial-With-R.html#6.1\%20Make\%20a\%20time\%20series\%20plot\%20(using\%20ggfortify)).

## Creating a project in RStudio

It is convenient to create an R project for each assignment that you are working on.
A project is basically a folder that stores all files related to the assignment.

You can create a project as follows:

* Open RStudio and click on "Project: (None)" in the top right corner.
* Open the dropdown window and click on "New Project...."
* In the popup window select "New Directory", then "New Project."
* Choose a sensible name for your project and enter it as the Directory Name. You can either use the default file path or change it by clicking "Browse..." next to "Create project as a subdirectory of:."
* Finally, click on "Create project."

After a project is created, there are two easy ways of accessing it. 
You can either use the same dropdown window in the top right corner of RStudio that you used to create the project, and click on the name of the project there, or you can find the project folder within your files and click on the file with the .Rproj extension.

## Installing packages

Most of the time when you work with R, you'll need to use functions or data from packages next to the base R functions that are automatically loaded when you open R.

One such package that you should almost always load when working with R is `tidyverse`, which is a collection of packages that allow clean workflows of data import, cleaning and manipulation in R.
The `ggplot2` package that you can use to create figures is one of these included packages.

You need to install each package only once, but you need to load them every time you open and use R.
It is good practice to load the package on the top of each script.
This is how you would install and load `tidyverse` (and any other package).

```{r, eval = FALSE}
install.packages("tidyverse") # install the package
```
```{r}
library(tidyverse) # load the package
```

## Supporting materials

You'll work on different assignments in different courses, but some general resources for working with R can be helpful in a wide range of assignments.
Below are some helpful links in case you would like to complement the Data Center workshops and tutorials (which you can find on the individual course pages).

### The basics of R

* [Cheat sheet for importing data](https://github.com/rstudio/cheatsheets/blob/main/data-import.pdf)
* [Cheat sheet for data manipulation](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf) (including `filter()`, `select()` and the pipe operator(`%>%`))
* [Cheat sheet for data tidying](https://github.com/rstudio/cheatsheets/blob/main/tidyr.pdf) (including reshaping and pivoting)
* [Getting started with Data in R](https://moderndive.netlify.app/1-getting-started.html#getting-started) (covers: using R and RStudio, basic terminology, introduction to packages)
* [R objects](https://rstudio-education.github.io/hopr/r-objects.html)
* [Advice on using projects](https://r4ds.had.co.nz/workflow-projects.html)
* [Full tutorial on R language basics (slides)](https://raw.githack.com/uo-ec607/lectures/master/04-rlang/04-rlang.html#1)
* [Full tutorial on data cleaning and wrangling (slides)](https://raw.githack.com/uo-ec607/lectures/master/05-tidyverse/05-tidyverse.html#1)
* [More explanation on the pipe operator](https://r4ds.had.co.nz/pipes.html)
* [Selecting and filtering data](https://mdsr-book.github.io/mdsr2e/ch-dataI.html#sec:pipe)
* [Conversions between long and wide format](https://mdsr-book.github.io/mdsr2e/ch-dataII.html#data-verbs-for-converting-wide-to-narrow-and-vice-versa)
* [Creating a project in RStudio (video)](https://www.youtube.com/watch?v=WyrJmJWgPiU)
* [Packages in R (video)](https://www.youtube.com/watch?v=v6VygIgvoZU&t=1s)
* [Loading a data file (video)](https://www.youtube.com/watch?v=2MVolYETR5Q)
* [Data manipulation (video)](https://www.youtube.com/watch?v=Zc_ufg4uW4U)

### Help with `ggplot`

* [Cheat sheet for `ggplot`](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf)
* [The R Graph Gallery (example plots, including full R scripts)](https://www.r-graph-gallery.com/)
* [More details on data visualization](https://mdsr-book.github.io/mdsr2e/ch-vizII.html#a-grammar-for-data-graphics)
* [Saving plots with `ggsave()`](https://www.tutorialgateway.org/save-r-ggplot-using-ggsave/)
* [Basic `ggplot` tutorial (video)](https://www.youtube.com/watch?v=hr2X7rmkprM)
* [Customizing plots (video)](https://www.youtube.com/watch?v=1GmQ5BdAhG4)

### Importing data

* [Cheat sheet for data import](https://github.com/rstudio/cheatsheets/blob/main/data-import.pdf)
* [More on `read_csv()`](https://uomresearchit.github.io/r-tidyverse-intro/03-loading-data-into-R/)
* [Import with `haven`](https://tutorials.methodsconsultants.com/posts/reading-sas-spss-or-stata-files-into-r-using-haven/)
* [Import with `readr` and `readxl` (video)](https://www.youtube.com/watch?v=366N1K1ja8g)
* [Import using `haven` (video)](https://www.youtube.com/watch?v=egFoDj6H9Ts)

### Data cleaning

* [Filter, select, mutate and summarize data](https://rpubs.com/AnalystNyamu/507453)
* [More on `mutate()`](https://www.sharpsightlabs.com/blog/mutate-in-r/)
* [Summarizing data by groups](https://r-graphics.org/recipe-dataprep-summarize)
* [List of common useful functions within `summarize()`](https://www.r-bloggers.com/2021/06/summarize-in-r-data-summarization-in-r/)
* [More on `rename()`](https://www.datasciencemadesimple.com/rename-the-column-name-in-r-using-dplyr/)
* [How to handle missing data](https://universeofdatascience.com/how-to-handle-missing-values-in-r/)
* [Select, filter, mutate and summarize data (video)](https://www.youtube.com/watch?v=nRtp7wSEtJA)
* [Summarizing by groups (video)](https://www.youtube.com/watch?v=zAlbrPozMHI)
* [How to handle missing values (video)](https://www.youtube.com/watch?v=O_gPPrezk5o)

### Merging data

* [`bind_rows()` and join functions visualized](https://www.r4epi.com/working-with-multiple-data-frames.html)
* [More on join functions](https://www.datasciencemadesimple.com/join-in-r-merge-in-r/)
* [Examples of using joins](https://statisticsglobe.com/r-dplyr-join-inner-left-right-full-semi-anti)
* [More on join functions (video)](https://www.youtube.com/watch?v=Yg-pNqzDuN4)
