---
title: "SSCECON207:<br>The European Debt Crisis"
date: "Spring 2023 (June)<br>Last updated: `r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, error = FALSE)
```
  
```{r}
library(tidyverse)
library(patchwork)

gdp <- read_csv("https://raw.githubusercontent.com/ucrdatacenter/projects/main/SSCECON207/2023h1/EU2009data/gdp_oecd.csv") %>% 
  filter(MEASURE == "USD_CAP")
unem <- read_csv("https://raw.githubusercontent.com/ucrdatacenter/projects/main/SSCECON207/2023h1/EU2009data/unem_oecd.csv") %>% 
  filter(FREQUENCY == "A", SUBJECT == "TOT") %>% 
  mutate(TIME = as.numeric(TIME))
infl <- read_csv("https://raw.githubusercontent.com/ucrdatacenter/projects/main/SSCECON207/2023h1/EU2009data/infl_oecd.csv") %>% 
  filter(FREQUENCY == "A", MEASURE == "AGRWTH", SUBJECT == "TOT") %>% 
  mutate(TIME = as.numeric(TIME))
irate <- read_csv("https://raw.githubusercontent.com/ucrdatacenter/projects/main/SSCECON207/2023h1/EU2009data/i_oecd.csv") %>% 
  filter(FREQUENCY == "A") %>% 
  mutate(TIME = as.numeric(TIME))
budg <- read_csv("https://raw.githubusercontent.com/ucrdatacenter/projects/main/SSCECON207/2023h1/EU2009data/budg_oecd.csv")
debt <- read_csv("https://raw.githubusercontent.com/ucrdatacenter/projects/main/SSCECON207/2023h1/EU2009data/debt_oecd.csv")

eu <- c("AUT", "BEL", "BGR", "CYP", "CZE", "DNK", "EST", "FIN", "FRA", "DEU", "GBR", "GRC", "HUN", "IRL", "ITA", "LVA", "LTU", "LUX", "MLT", "NLD", "POL", "PRT", "ROU", "SVK", "SVN", "ESP", "SWE", "OECDE", "EA19")
```
  

# Introduction

This case study looks at the European Debt Crisis between 2009-2012 by combining insights from the IS-LM-PC model with empirical analysis.
The Great Recession, which started in the US with the burst of the housing bubble, also had a large impact on all European economies.
In addition, the EU faced a unique situation due to its high levels of real and financial integration.
Therefore many European countries -- Greece being the most well-known -- ended up with unmanageable levels of government budget deficit and government debt.
Policy interventions helped somewhat, but the EU nevertheless suffered the consequences of the crisis for several years.

In the following we examine why the debt crisis unfolded, the effectiveness of policy interventions by looking at the evolution of macroeconomic indicators.
At the end of the case study we estimate the Phillips curve for the Euro Area between 2000-2015 to see how well the theoretical relationship between inflation and unemployment held in the period of the debt crisis.

***Note:***  
*The code used to create the figures used in this case study is available on [Github](https://github.com/ucrdatacenter/projects/blob/main/SSCECON207/2023h1/EU2009code.R).*

# Background

## Macro indicators before the crisis

In the years before the crisis, the economies of the European Union (EU) were doing quite well.

As the figure below shows, the average GDP growth between 2000-2007 was above 3% in all EU economies, and some countries (e.g. the Baltics, Southern countries, Ireland) showed particularly strong growth in the period.
The time-series plot showing the annual growth of the Euro Area since 2000 also shows strong annual growth rates between 2-8%.

```{r, fig.height=7}
p1 <- gdp %>% 
  filter(TIME %in% 1999:2007, LOCATION %in% eu) %>% 
  arrange(TIME) %>% 
  group_by(LOCATION) %>% 
  mutate(growth = (Value-lag(Value))/lag(Value)*100) %>% 
  summarize(Value = mean(growth, na.rm = TRUE)) %>% 
  ggplot(aes(LOCATION, Value)) +
  geom_col() +
  xlab(NULL) + ylab("Average GDP growth 2000-2007 (%)") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90))

p2 <- gdp %>% 
  filter(TIME %in% 1999:2007, LOCATION == "EA19") %>% 
  mutate(growth = (Value-lag(Value))/lag(Value)*100) %>% 
  ggplot(aes(TIME, growth)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(limits = c(2000, 2007)) +
  xlab("Year") + ylab("Euro Area GDP growth (%)") +
  theme_light()

p1 + p2 + plot_layout(ncol = 1)
```

While the European unemployment rate is generally higher than the US unemployment rate, average unemployment in the years before the crisis was also not a major issue: it went above 10% only in some of the former Eastern Block countries, as a consequence of their still relatively recent transition.
The Euro Area in particular was doing well in terms of reducing unemployment: in the years before the crisis, unemployment was falling slowly but consistently, reaching below 8% by 2007.

```{r, fig.height=7}
p1 <- unem %>% 
  filter(TIME %in% 2000:2007, LOCATION %in% eu) %>% 
  group_by(LOCATION) %>% 
  summarize(Value = mean(Value, na.rm = TRUE)) %>% 
  ggplot(aes(LOCATION, Value)) +
  geom_col() +
  xlab(NULL) + ylab("Average unemployment rate 2000-2007 (%)") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90))

p2 <- unem %>% 
  filter(TIME %in% 2000:2007, LOCATION == "EA19") %>% 
  ggplot(aes(TIME, Value)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(limits = c(2000, 2007)) +
  xlab("Year") + ylab("Euro Area unemployment rate (%)") +
  theme_light()

p1 + p2 + plot_layout(ncol = 1)
```

## European integration

Euro Area needs to coordinate economic policy, which means that the European Central Bank (ECB) sets the same interest rate for all Euro Area countries.
Therefore, in the case of a crisis, the available monetary policy tools are limited, because different Euro Area countries may be impacted differently, requiring different policy interventions, but the ECB can only choose one interest rate.

In addition, all EU countries are part of the Stability & Growth Pact (SGP), which limits government deficit to 3% of GDP and public debt to 60% of GDP.
While this pact allows for temporary deviations in a crisis, it still places some limits on fiscal policy, especially in the recovery stage after a crisis.
Other than the SGP, fiscal policy is largely in the hands of national governments, uncoordinated between EU countries.

These limitations on monetary policy and coordination issues in fiscal policy were not very problematic in the stable economic conditions before the crisis, but you will see that they affected the policy choices during and after the crisis.
Already in the patterns in GDP and unemployment above, you can see that even within the Euro Area, countries had different growth and unemployment rates.
Recall that faster growing countries would like higher interest rates to keep inflation low, while slower growing countries prefer lower interest rates and expansionary fiscal policy to stimulate growth.

# The debt crisis

While the US debt crisis is considered to be the initial shock leading to the European debt crisis, Europe had plenty of existing problems that made countries more vulnerable to shocks.
For example, the UK, Ireland and Spain had housing bubbles similar to the US, while the governments Greece and Italy were accumulating a lot of debt to finance their budget deficits already in the years prior to the crisis.
The main reason for these patterns was the ECB's monetary policy: 
since the ECB could only set one interest rate in the Euro Area, the faster-growing, mostly Southern countries had lower than optimal interest rates, which incentivized them to borrow money. 
In countries like Ireland and Spain these interest rates primarily affected private borrowing, while in Italy and Greece the rise in government borrowing was notable, leading to increasing budget deficits and government debt levels.

When the US housing bubble burst, the effects quickly spread over to Europe.
The European housing bubbles also burst, and lending became more risky, leading to higher risk premia and therefore more expensive borrowing.
The increasing costs of borrowing made the high debt levels of countries like Greece unsustainable.

The following figure presents the disparity between EU countries in their government budget deficits and government debts (both as percentage of GDP) in 2008 and 2009.
Clearly, both the overall levels of the indicators and the differences between countries only got worse as the Great Recession unfolded.

```{r}
bind_rows(budg, debt) %>% 
  filter(TIME %in% 2008:2009, LOCATION %in% eu) %>% 
  mutate(INDICATOR = ifelse(INDICATOR == "GGDEBT", "Government debt", "Budget deficit")) %>% 
  ggplot(aes(LOCATION, Value)) +
  geom_col() +
  facet_grid(~INDICATOR~TIME, scales = "free_y") +
  xlab(NULL) + ylab("% of GDP in 2008") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))
```

Greece had the highest government debt to GDP ratio (over 100%) and the largest budget deficit (10% of Greek GDP) out of all EU countries in both years.
Therefore it is not surprising that Greece's debt became the most closely associated with the debt crisis: it was the first country to be unable to pay back its debt.

Ireland, Portugal, Italy and Spain were also at risk of defaulting on their debt, thereby risking the financial stability of the entire Euro Area.
There was a need for effective policy interventions from the ECB, the EU, and national governments.

# Policy responses

## Monetary policy

The following figure plots the interest rate set by the ECB in the Euro Area.
Before the crisis, it maintained relatively high interest rates to keep inflation low in the strong economies of the Euro Area.
Once the crisis hit Europe, interest rates were dropped steeply as part of an expansionary monetary policy (together with multiple bailout packages for the countries most affected by the debt crisis).

In terms of the IS-LM-PC model, the ECB reacted to output falling below potential as a result of the Great Recession, and it shifted the LM curve down to increase output.
However, in doing so, the ECB still had to take into account both the different macroeconomic states within the Euro Area and the zero lower bound on interest rates.
So monetary policy alone, despite the intensity of the intervention, was not enough for recovery. 

```{r}
irate %>% 
  filter(TIME %in% 2000:2019, LOCATION == "EA19") %>% 
  ggplot(aes(TIME, Value)) +
  geom_point() +
  geom_line() +
  xlab(NULL) + ylab("Euro Area interest rate (%)") +
  theme_light()
```


## Fiscal policy

Before the crisis, coordinating fiscal policy could have alleviated many of the issues the EU faced.
Since monetary policy was centralized, so countries could not use the interest rate (LM curve) to bring output to potential, fast-growing countries could have used fiscal contraction, and slow-growing countries fiscal expansion.
In theory, doing so would over time bring all outputs to potential, therefore eliminating the discrepancy within the Euro Area.

Once the Great Recession's effects reached Europe, such policy coordination was no longer feasible, as all countries needed fiscal expansion to combat the recession.
On the other hand, countries receiving bailout packages from the EU or the IMF were forced to adopt austerity measures (i.e. fiscal contraction) to reduce the budget deficit -- the exact opposite to what the countries needed to recover from the recession.

# Recovery and results

## GDP growth and unemployment

If we look at the Euro Area as a whole -- as in the figure below -- the crisis in terms of output growth does not seem extreme.
While output growth was negative in 2009, the Euro Area had stable positive output growth from 2010 onward.
The unemployment rate looks worse: it only peaked in 2013 at 12%, before starting to decrease again.

```{r}
bind_rows(gdp, unem) %>% 
  filter(LOCATION == "EA19") %>% 
  arrange(TIME) %>% 
  group_by(INDICATOR) %>% 
  mutate(Value = ifelse(INDICATOR == "GDP", 
                        (Value-lag(Value))/lag(Value)*100, Value),
         INDICATOR = ifelse(INDICATOR == "GDP", 
                            "GPD growth", "Unemployment rate")) %>% 
  filter(TIME %in% 2000:2019) %>% 
  ggplot(aes(TIME, Value)) +
  geom_point() +
  geom_line() +
  facet_wrap(~INDICATOR, ncol = 1, scales = "free_y") +
  xlab(NULL) + ylab("Euro Area (%)") +
  theme_bw()
  
```

However, looking at the Euro Area as a whole does not show the full picture.
Below you can see the same figure of GDP growth and unemployment for some of the countries most affected by the crisis: Greece, Ireland, Spain and Portugal (with Germany added for comparison).

```{r}
bind_rows(gdp, unem) %>% 
  filter(LOCATION %in% c("DEU", "ESP", "GRC", "IRL", "PRT")) %>% 
  arrange(TIME) %>% 
  group_by(INDICATOR, LOCATION) %>% 
  mutate(Value = ifelse(INDICATOR == "GDP", 
                        (Value-lag(Value))/lag(Value)*100, Value),
         INDICATOR = ifelse(INDICATOR == "GDP", 
                            "GPD growth", "Unemployment rate")) %>% 
  filter(TIME %in% 2000:2019) %>% 
  ggplot(aes(TIME, Value, color = LOCATION)) +
  geom_point() +
  geom_line() +
  facet_wrap(~INDICATOR, ncol = 1, scales = "free_y") +
  scale_x_continuous(limits = c(2000, 2018)) +
  xlab(NULL) + ylab("%") +
  theme_bw()
```

Greece endured three years of negative output growth (close to -10%), while Ireland had two years of negative growth.
Both Greece's and Spain's unemployment rates peaked above 25%, significantly above the Euro Area average.
Meanwhile, the German unemployment rate was completely unaffected by the crisis, and German economic growth already showed a strong recovery in 2010: the crisis clearly produced vastly different macroeconomic effects between countries.

These differences are explained by the policies used in different countries.

Once the crisis hit, Germany benefited from the expansionary monetary policy of the ECB and its national government's fiscal expansion.
The Great Recession caused a leftward shift of the German IS curve, but the swift downward shift of the LM curve via the low interest rate and the rightward shift of the IS curve via fiscal expansion quickly returned output growth to its pre-crisis level.
As the unemployment rate takes time to adjust to a recession, a quick recovery can prevent a sustained increase in the unemployment rate.

On the other hand, countries that had to adopt fiscal contraction due to IMF-imposed austerity measures recovered slowly, as the fiscal policy shifted the IS curve even further left.
The resulting shift of the IS curve was so large that the ECB's monetary expansion did not push the LM curve down sufficiently: output growth did not return to its pre-crisis level for years.
The fall in output also impacted the unemployment rate for several years after the start of the recession: these countries had a long and painful recovery.

## The Phillips curve in the Euro Area

Given the large fluctuations in the unemployment rate before, during and after the crisis in many European countries, it is interesting to look at whether the Phillips curve relationship between inflation and unemployment held in this period in the EU.
The figure below shows scatterplots of the inflation rate and the unemployment rate between 2000-2016 in some of the countries most affected by the debt crisis.
The blue lines are the estimated Phillips curves per country.

```{r}
bind_rows(infl, unem) %>% 
  filter(LOCATION %in% c("GRC", "ESP", "IRL", "PRT"), TIME %in% 2000:2016) %>% 
  select(LOCATION, INDICATOR, TIME, Value) %>% 
  pivot_wider(names_from = INDICATOR, values_from = Value) %>% 
  ggplot(aes(HUR, CPI)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Unemployment rate") + ylab("Inflation rate") +
  facet_wrap(~LOCATION, scales = "free", dir = "v") +
  theme_bw()
```

While the relationships are not very strong, all countries exhibited a negative relationship between inflation and unemployment: the high unemployment years post-crisis corresponded to almost zero inflation or even deflation.
