---
layout: page
title: "Data visualization: distributions"
date: "Last updated: `r Sys.Date()`"
output:
  md_document:
    variant: gfm
    preserve_yaml: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

**This tutorial is not ready yet. Please come back later.**

# Introduction

Before you proceed, make sure you're familiar with the logic of `ggplot`, as explained in our [introduction to `ggplot` tutorial](../r_ggplot_intro).

We'll use the `diamonds` dataset that comes pre-loaded with `tidyverse` to demonstrate how to visualize the distribution of a variable with `ggplot2`, so let's load `tidyverse` and have a look at the dataset:

```{r}
# load tidyverse
library(tidyverse)

# add diamonds to the environment
data(diamonds)
```

# Univariate plots

You can plot a single continuous variable with a histogram, a density plot, or a boxplot.
Other than the name of the dataset and the variable that you want to plot, no additional arguments need to be specified; but you can customize the plot by adding arguments to the `geom_` functions.

```{r}
# basic histogram of price
ggplot(diamonds, aes(x = price)) +
  geom_histogram()

# custom binwidth or bins determine the number of bins
# with binwidth = 1000, each bin is $1,000 wide
# color affects the border; fill affects the inside
ggplot(diamonds, aes(x = price)) +
  geom_histogram(binwidth = 1000, color = "black", fill = "lightblue")

# density plot
# alpha adjusts the transparency of the fill
ggplot(diamonds, aes(x = price)) +
  geom_density(fill = "lightblue", alpha = 0.5)

# boxplot
ggplot(diamonds, aes(x = price)) +
  geom_boxplot()
```

To compare the frequencies of discrete variables, you can use a bar plot.

```{r}
ggplot(diamonds, aes(x = cut)) +
  geom_bar()
```

Whenever you make a plot, make sure to use clear labels and titles with the `labs()` function to make your visualization easy to understand.

```{r}
ggplot(diamonds, aes(x = cut)) +
  geom_bar() +
  labs(title = "Frequencies of diamond cuts",
       x = "Cut",
       y = "Number of diamonds")
```

To learn more about other geoms and customization options, have a look at our [advanced visualization tutorial](../r_adv_ggplot) and [additional resources](../links#Help with `ggplot`).

## Video tutorial TBA