---
layout: page
title: "SSCPOLI302:<br> Analysis of MEPs (draft)"
subtitle: "Fall 2024"
date: "Last updated: `r Sys.Date()`"
output:
  md_document:
    variant: gfm
    preserve_yaml: true
    toc: true
---

# Loading libraries and data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

```{r}
# install.packages("tidyverse")
# install.packages("rio")
library(tidyverse)
library(rio)

meetings <- import("https://github.com/ucrdatacenter/projects/raw/main/SSCPOLI302/2024h2/Meetings_compl.xlsx", setclass = "tbl_df") |> 
  mutate(meeting_date = as.Date(meeting_date))
```

# Describing the MEPs in the data

List and count the total number of MEPs in the data.

```{r}
# get all MEP names, MEP ID's, and nationalities
MEPs <- meetings |> 
  distinct(member_name, member_id, country)
  
head(MEPs) # show the first few rows
nrow(MEPs) # number of MEPs in the data
```

Not all of the 705 MEPs of term 2019-2024 are represented in the data.

# Differences between member states

```{r}
# count the number of MEPs per country
MEPs |>
  count(country, sort = TRUE)
```

```{r}
# count the number of meetings per country
meetings |> 
  count(country, sort = TRUE)
```

We can show differences between countries by plotting the number of meetings per country on a bar chart.
  
```{r}
# plot the number of meetings per country
meetings |> 
  count(country, sort = TRUE) |> 
  ggplot() +
  geom_col(aes(x = n, y = country)) +
  labs(title = "Number of meetings per country between 2019-2024", 
       x = "Country", 
       y = "Number of meetings")
```

We can improve the display by arranging the bars based on the number of meetings with the `reorder()` function.

```{r}
# plot the number of meetings per country
meetings |> 
  count(country, sort = TRUE) |> 
  ggplot() +
  geom_col(aes(x = n, y = reorder(country, n))) +
  labs(title = "Number of meetings per country between 2019-2024", 
       x = "Country", 
       y = "Number of meetings")
```

In this figure, the largest countries have the most meetings, which is expected given that they have more MEPs.
To account for this difference, we can instead plot the average number of meetings per MEP in each country, by dividing the number of meetings by the number of MEPs.

```{r}
# average number of meetings per MEP in each country
countries <- meetings |> 
  group_by(country) |>
  summarize(n_meetings = n(),
            n_meps = n_distinct(member_id),
            avg_meetings = n_meetings / n_meps)
```

To visualize differences between countries in the average number of meetings of their MEPs, we can present the data on a bar chart.

```{r}
# plot the average number of meetings per MEP for each country
countries |> 
  ggplot() +
  geom_col(aes(x = avg_meetings, y = reorder(country, avg_meetings))) +
  labs(title = "Average number of meetings per MEP between 2019-2024", 
       x = "Country", 
       y = "Average number of meetings per MEP")
```

This figure tells us more about the activity of MEPs in each country, as it accounts for the number of MEPs in each country.
We can see that MEPs from Finland are particularly active.
On the other hand, Cyprus not only has the fewest MEPs but also the fewest meetings per MEP.

# Meeting timelines

Some countries have more MEPs than their EP seats. 
This means that an EP member did not finish their term and was replaced by another, such as the case of Liesje Schreinemacher.

We can plot the distribution of meetings over time with a histogram.
A histogram aggregates the data into bins and counts the number of observations in each bin.
In this case, we set each bin to be 30 days wide.

```{r}
# meeting dates of Liesje Schreinemacher
meetings |>
  filter(member_name == "SCHREINEMACHER Liesje") |> 
  ggplot() +
  geom_histogram(aes(x = meeting_date), binwidth = 30) +
  labs(title = "Meeting activity of Liesje Schreinemacher during her
         MEP career", 
       x = "Date", 
       y = "Density")
```

We can also compare the density of Liesje Schreinemacher's meetings to the overall meeting density of the EP.
We will do this by changing a few things within the code.

First, we take the `meetings` dataset and define a new variable `MEP` that indicates whether the MEP is Liesje Schreinemacher or not.
Then, we plot the histogram of the density of meetings for both Liesje Schreinemacher and the other MEPs.
We set the fill of the histogram to be different depending on whether the MEP is Liesje Schreinemacher or not.

```{r}
meetings |> 
  mutate(MEP = ifelse(member_name == "SCHREINEMACHER Liesje",
                      "Liesje Schreinemacher",
                      "Other MEPs")) |>
  ggplot() +
  geom_histogram(aes(x = meeting_date, fill = MEP), binwidth = 30)
```

By default the histogram shows the count of meetings.
Compared to the sum all MEPs, Liesje Schreinemacher had so few meetings that her histogram is barely visible.
By specifying the `y` aesthetic to be a rescaled version of the count (so that the highest value within a group is 1), we override this default behavior and make the comparison more informative.
We accomplish this by specifying `after_stat(ncount)` as the `y` aesthetic.
In addition, we can put the bars of the groups next to each other instead of stacking them by specifying `position = "dodge"`.

```{r}
meetings |> 
  mutate(MEP = ifelse(member_name == "SCHREINEMACHER Liesje",
                      "Liesje Schreinemacher",
                      "Other MEPs")) |>
  ggplot() +
  geom_histogram(aes(x = meeting_date, y = after_stat(ncount), fill = MEP),
                 binwidth = 30, position = "dodge") +
  labs(title = "Meeting distribtion of Liesje Schreinemacher compared to
         the rest European Parliament between 2019-2024",
       x = "Date", 
       y = "Distribution of meetings")
```

As we can see, Liesje Schreinemacher's activity stops about midway through 2022, which is when she exited the EP.

# Differences between political groups

First, we can compare the number of meetings from each political group who declared meetings compared to the total number of seats of each group.
Calculating the average number of meetings per seat in the political groups helps us see which group has held the most meetings on average while accounting for their size.
We use seats because only one MEP can occupy a seat at a time, therefore this definition accounts for groups that saw many MEP replacements and thus have more MEPs than seats in their data.

```{r}
# define the number of seats per group (as of May 2024)
seats <- tribble(
  ~political_group, ~seats,
  "EPP", 176,
  "S&D", 139,
  "Renew", 102,
  "Greens", 72,
  "ECR", 69,
  "Non-attached", 61,
  "ID", 49,
  "the Left", 37)

# count the number of meetings per political group in the meeting data
meetings_per_seat <- meetings |> 
  count(political_group) |> 
  # merge with the number of seats per group
  left_join(seats, by = "political_group") |> 
  # calculate the number of meetings per seat
  mutate(meetings_per_seat = n / seats)
```

We can compare visualizations for the number of MEPs per political group and the average number of meetings per seat in each political group using bar charts.

```{r}
# number of MEP seats per political group
meetings_per_seat |> 
  ggplot() +
  geom_col(aes(x = political_group, y = seats)) +
  labs(title = "Number of MEPs per political group in May 2024", 
       x = "Political group", 
       y = "Number of MEPs")
```

To improve the quality of the graph, we can again organize the bars in ascending order of the number of MEPs.
In addition, we can specify the colors of the bars to match those of their political groups.
We define these colors in a separate vector and use the `scale_fill_manual()` function to apply them to the plot.

```{r}
# define political group colors
colors <- c("EPP" = "deepskyblue", 
            "S&D" = "red",
            "Renew" = "gold", 
            "Greens" = "forestgreen",
            "ECR" = "deepskyblue4", 
            "ID" = "darkblue", 
            "the Left" = "darkred", 
            "Non-attached" = "grey")

# number of MEP seats per political group
meetings_per_seat |> 
  ggplot() +
  geom_col(aes(x = reorder(political_group, seats), y = seats, 
               fill = political_group),
           # color the borders of the bars black and hide the legend
           color = "black", show.legend = FALSE) +
  labs(title = "Number of MEPs per political group in May 2024", 
       x = "Political group", 
       y = "Number of MEPs") +
  scale_fill_manual(values = colors)
```

Now compare the previous plot to the average number of meetings per seat in each political group.

```{r}
# average number of meetings per seat per political group
meetings_per_seat |> 
  ggplot() +
  geom_col(aes(x = reorder(political_group, meetings_per_seat), 
               y = meetings_per_seat, fill = political_group),
           color = "black", show.legend = FALSE) +
  labs(title = "Average number of reported meetings per seat", 
       x = "Political group", 
       y = "Number of meetings per seat") +
  scale_fill_manual(values = colors)
```

# Role distributions

Meetings also differ based on the roles of the MEP in the meeting.
First, let's compare the number of meetings for each role.

```{r}
meetings |> 
  count(member_capacity, sort = TRUE)
```

We can combine this information with the political group of the MEP to see which combination of political groups and roles are most common.

```{r}
meetings |> 
  count(political_group, member_capacity, sort = TRUE)
```

This is a lot of information as a table, so let's use a bar chart to show the distribution of political groups within each role.

```{r}
meetings |> 
  count(political_group, member_capacity) |> 
  ggplot() +
  geom_col(aes(x = n, y = member_capacity, fill = political_group)) +
  labs(title = "Meeting frequency per role and political group", 
       x = "Number of meetings",
       y = "Role",
       fill = "Political group") +
  scale_fill_manual(values = colors)
```

*Unclear what's happening here, to be discussed*

```{r}
#Optional extra; try to figure out which roles have linked a procedure reference.
#This is imporant for full transparency about the legislative dossier their
#meeting relates to.


#Computing which roles have added a reference with a case_when function.
#Afterwards computing what percentage of each role then referenced a file.
#Finally ranking the results by % within "yes" or "no" reference-groups.

Role_reference <-
    meetings |>
    group_by(member_capacity) |>
    mutate(reference = if_else(is.na(procedure_reference), "no", "yes")) |>
    group_by(member_capacity, reference) |>
    summarize(nr = n()) |>
    mutate(role_total = sum(nr),
           reference_percentage = ((nr/role_total)*100)) |>
    group_by(reference) |>
    arrange(desc(reference_percentage), .by_group = TRUE)


```
