---
layout: page
title: "SCIBIOM303 Workshop 2: Temperature and Heatstroke-related Ambulance Dispatches"
subtitle: "Spring 2025"
date: "Last updated: `r Sys.Date()`"
output:
  md_document:
    variant: gfm
    preserve_yaml: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, error=FALSE)
```

# Introduction
In this encounter, we will investigate and analyze the relationship between temperature and heatstroke-related ambulance dispatches using data from Japan. This data offers information about the maximum temperature and relative humidity measured per day in all 47 prefectures from 2015 to 2019.

# Loading packages

First, we load the `tidyverse` and `rio` packages.
Make sure that all packages are installed before running the code.

```{r}
# install.packages("tidyverse")
# install.packages("rio")
library(tidyverse)
library(rio)
```

# Importing data sets

We import the data sets for the encounter.

```{r}
hsad <- import("https://github.com/ucrdatacenter/projects/raw/refs/heads/main/SCIBIOM303/2025h1/data/HSAD.csv")
temp <- import("https://github.com/ucrdatacenter/projects/raw/refs/heads/main/SCIBIOM303/2025h1/data/temperature.csv")
```

# Merging the two data sets

We merge the two data sets together. We can do this by using the function `left_join()`, which enables us to merge all the columns from both data sets together. We join them by the columns 'Date' and 'Prefecturename' which are the same in both data sets to avoid having double values. Then, to make it easier to work with the columns, we remove a column called 'Prefecture', as it only refers to the ID number of each prefecture. We then rename the column 'Prefecturename' to 'Prefecture'.

```{r}
merged <- left_join(temp, hsad, by = c("Date", "Prefecturename")) |> 
  select(-Prefecture) |> 
  rename("Prefecture" = Prefecturename)
```

# Choosing a prefecture and a year

Now we can investigate the relationship between maximum temperature and heatstroke-related ambulance dispatches (HSAD) per day in one prefecture in one year. We can choose Hiroshima, and the year 2015. We also change the variable 'Date' from character to a proper date format to better visualize it later. We do this by creating a pipeline using the function filter() to filter for the selected parameters, and mutate() to change the variable 'Date'.

```{r}
hiroshima_2015 <- merged |> 
  filter(Prefecture == "Hiroshima") |> 
  filter(Year == 2015) |> 
  mutate(Date = ymd(Date)) 
```

# Creating a plot with Tempmax and HSAD

Now we can create a scatter plot by using the function geom_point() to showcase the relationship between the maximum temperature (Tempmax) and heatstroke-related ambulance dispatches (HSAD). We also add geom_smooth() to showcase a trend line. We can add names of the axes and a title of the graph.

```{r}
ggplot(hiroshima_2015, aes(x = Tempmax, y = HSAD)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(title = "Temperature and Heatstroke-related Ambulance Dispatches in Hiroshima in 2015",
       x = "Maximum Temperature",
       y = "Heatstroke-related Ambulance Dispatches") +
  theme_minimal()
```

We can see that HSAD is consistently near 0 degrees Celsius between temperatures 20-25, and after 30 degrees Celsius, it increases steeply, almost in a linear way. This is as expected, as being in environments with higher temperature leads to a rise in core body temperature, which can lead to the development of heatstroke ([source](https://www.mayoclinic.org/diseases-conditions/heat-stroke/symptoms-causes/syc-20353581)). 

# Creating a plot with Tempmax and Date

Now we can create a scatter plot to showcase the trends in the maximum temperature (Tempmax) over time (Date) in Hiroshima in 2015.

```{r}
ggplot(hiroshima_2015, aes(x = Date, y = Tempmax)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(title = "Temperature Over Time in Hiroshima in 2015",
       x = "Date",
       y = "Maximum Temperature") +
  theme_minimal()
```

We can see that the temperature starts rising in June and peaks in August (>35 degrees Celsius), and then declines again until October. This is what we would expect, as this is the typical pattern in temperature in Hiroshima across these months ([source](https://www.climatestotravel.com/climate/japan/hiroshima)).

# Linear regression analysis

We can perform a linear regression analysis to analyze if there is a relationship between maximum temperature (Tempmax) and heatstroke-related ambulance dispatches (HSAD) in Hiroshima in 2015. We can do this by simply using the functions lm(), to perform the linear regression analysis, and then summary(), to summarize and show the results of the analysis.

```{r}
lr <- lm(HSAD ~ Tempmax, data = hiroshima_2015)
summary(lr)
```

The results show that R-squared = 0.6193; this means that about 61.93% of the variability in HSAD is explained by Tempmax. This indicates a moderate to strong positive relationship between temperature and heat stroke-related ambulance dispatches.

# Comparing two prefectures

Now we can compare the relationship in two prefectures in the same year. We can choose Hiroshima and Kyoto in 2015. We can do this by creating a pipeline using our merged data set. We filter for the two prefectures and the year, and then we make a line graph using geom_smooth().

```{r}
merged |> 
  filter(Prefecture == "Kyoto"| Prefecture == "Hiroshima") |> 
  filter(Year == 2015) |> 
  ggplot() +
  geom_smooth(aes(x = Tempmax, y = HSAD, color = Prefecture), se = FALSE) +
  labs(title = "Temperature and Heatstroke-related Ambulance Dispatches (2015)",
       x = "Maximum Temperature",
       y = "Heatstroke-related Ambulance Dispatches") +
  theme_minimal()
```

This graphs shows similar trends in the relationship between temperature and heatstroke-related ambulance dispatches in both prefectures. While specific studies comparing Hiroshima and Kyoto are limited, research suggests a consistent relationship between higher temperatures and increased heatstroke-related ambulance dispatches across various regions in Japan ([source](https://doi.org/10.3961/jpmph.2012.45.5.309)).

# Comparing the days of the week

Now we can investigate whether there is a difference between the number of dispatches during the week and during the weekend. We can use Hiroshima in 2015 as an example. We create a pipeline; we use the group_by() function to group rows by the day of the week variable (Dow). Next, within summarize(), which reduces multiple rows in each group to a single summary value, we apply mean() to calculate the average number of heatstroke-related ambulance dispatches for each day of the week. Then we can create a bar chart to show the average number of dispatches per day during the week.

```{r}
hiroshima_2015 |>
  group_by(Dow) |>
  summarize(HSAD = mean(HSAD)) |>
  ggplot(aes(x = Dow, y = HSAD)) +
  geom_col() +
  labs(title = "Heatstroke-related Ambulance Dispatches by Day of the Week (2015)",
       x = "Day of the week",
       y = "Heatstroke-related Ambulance Dispatches") +
  theme_minimal()
```

We can see that there are more cases of HSAD on days 1 and 7, which refer to Sunday and Saturday, respectively. These results are as expected, as more cases of heatstroke occur most commonly on weekends, especially on Sundays when people tend to take part in sporting or outdoor activities ([source](https://www.channelnewsasia.com/today/ground-up/heatstroke-heat-related-illnesses-dangers-risks-take-precautions-misconceptions-4647621)).

# Creating heat index

Now we can create a new variable called heat index. This measure is calculated from temperature and humidity. First we need to convert the temperature from Celsius to Fahrenheit and create a new variable called Temp_F; then we calculate the heat index in Fahrenheit (Heat_Index_F), and then we convert it to Celsius (Heat_Index_C). To do this, we use the function mutate() to create these new variables.

```{r}
hiroshima_lr_data <- hiroshima_2015 %>%
  mutate(
    Temp_F = Tempmax * 9 / 5 + 32,
    Heat_Index_F = -42.379 + 
      2.04901523 * Temp_F + 
      10.14333127 * Rhumave - 
      0.22475541 * Temp_F * Rhumave - 
      0.00683783 * Temp_F^2 - 
      0.05481717 * Rhumave^2 + 
      0.00122874 * Temp_F^2 * Rhumave + 
      0.00085282 * Temp_F * Rhumave^2 - 
      0.00000199 * Temp_F^2 * Rhumave^2,
    Heat_Index_C = (Heat_Index_F - 32) * 5 / 9)
```

# Creating a plot with Heat_Index_C and HSAD

Now we can create a plot to show the relationship between the heat index and HSAD. We use the same code as in the first plot, but we use the hiroshima_lr_data and instead of maximum temperature, we use the heat index in Celsius (Heat_Index_C) on the x-axis.

```{r}
ggplot(hiroshima_lr_data, aes(x = Heat_Index_C, y = HSAD)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(title = "Heat Index and Heatstroke-related Ambulance Dispatches in Hiroshima in 2015",
       x = "Heat Index",
       y = "Heatstroke-related Ambulance Dispatches") +
  theme_minimal()
```

We can see that heat index is consistently near 0 between temperatures 20-30, and after 30 degrees Celsius, it increases sharply in an almost linear way. This is as expected, as the heat index positively correlates with the rate of onset for heat stroke ([source](https://doi.org/10.1186/s40779-015-0056-z)).

# Linear regression analysis

We can perform a linear regression analysis to analyze if there is a relationship between the heat index (Heat_Index_C) and heatstroke-related ambulance dispatches (HSAD) in Hiroshima in 2015.

```{r}
lr2 <- lm(HSAD ~ Heat_Index_C, data = hiroshima_lr_data)
summary(lr2)
```

The results show that R-squared = 0.7222; this means that about 72.22% of the variability in HSAD is explained by the heat index. This indicates an even stronger positive relationship between the heat index and heat stroke-related ambulance dispatches.
