---
title: "Data Center Apprenticeship:\nEfficient workflows with GitHub and Quarto"
subtitle: "January 2025" 
date: "Last updated: `r Sys.Date()`"
output:
  md_document:
    variant: gfm
    preserve_yaml: true
    toc: true
---

# Introduction

In this tutorial, we will explore how to use GitHub and Quarto to streamline your workflows, improve collaboration, and create professional, reproducible documents. 

This tutorial is divided into two parts:

1. An introduction to using GitHub for version control and collaboration. GitHub is a version control platform that allows you to track changes, collaborate with others, and manage your projects efficiently. GitHub Desktop is a graphical user interface that simplifies working with Git repositories, making it more accessible to users.
2. An overview of creating and using Markdown documents with Quarto for documentation and reporting. Quarto is a tool for creating documents using Markdown, enabling you to generate reports, presentations, and analyses with ease. (If you are familiar with RMarkdown, Quarto is a very similar tool with additional features.)

# Working with GitHub using GitHub Desktop

## Introduction to GitHub

GitHub is a web-based platform designed for version control and collaboration. 
At its core, GitHub uses repositories (often abbreviated as "repos") to organize and store project files, including their full revision history. 
A repository acts as a central storage location where you can manage code, track changes, and collaborate with others.

You can create your own repositories, access any other public repository, and contribute to open-source projects.
You can also use GitHub to track issues, manage projects, and host documentation.
For example, R packages are often hosted on GitHub, making it easy for users to install and update packages directly from the platform, or view the source code.

GitHub Desktop is a graphical user interface that simplifies working with GitHub repositories. It eliminates the need to use the command line, making Git more accessible. In this section, we’ll explore the essential features of GitHub Desktop, including cloning and forking repositories, committing changes, pushing updates, and creating pull requests. These features help manage version control, ensure reproducibility, and facilitate collaboration.

In the following, we guide you through an example workflow using GitHub Desktop.
This example is similar to the process you might follow if contributing to the Data Center's projects.

Before we begin, make sure that you have created a [GitHub account](https://github.com/) and installed [GitHub Desktop](https://desktop.github.com/download/) on your computer.

## Forking a Repository

Forking creates your own copy of someone else’s repository, allowing you to make changes without affecting the original project. This is ideal when contributing to open-source projects or experimenting with new ideas.
It is also a good way of making sure that the original repository remains stable, as any changes to it need to be approved by the repository owner.

When working with the DC's repositories, you will typically fork the repository to your own GitHub account, make changes in your fork, and then create a pull request to propose changes to the original repository.
This method prevents version clashes if multiple contributors are working on the same project.

Steps of forking a repository:

1. Navigate to the repository page in your browser: `https://github.com/ucrdatacenter/projects`.
2. Click the `Fork` button at the top-right corner of the page.
3. Select your GitHub account as the destination for the fork.
4. Clone your forked repository to your local machine using GitHub Desktop (refer to the steps for cloning above).

Forking is especially beneficial for collaboration. It ensures you can test and develop features independently before proposing changes to the main project.

```r
# Placeholder for screenshot of forking process
```

---

## Cloning a repository

Cloning a repository is useful when you want to work on an existing project locally.
It is the most convenient way of working with files, as you can make changes while working normally in your local directory and with any software installed on your computer.

The forked repo can be your own project or one you’re contributing to. For example, if you’re working on a collaborative project like `ucrdatacenter/projects`, cloning allows you to download the repository to your local machine for modification.

Steps:

1. Open GitHub Desktop.
2. Go to `File > Clone Repository...`.
3. Select the `URL` tab and either choose the repository from the list or enter the repository URL: `https://github.com/ucrdatacenter/projects`.
4. Choose a local path where the repository will be saved and click `Clone`.

```r
# Placeholder for screenshot of cloning process
```

---

### Making Commits and Pushing Changes

Commits are snapshots of your changes that you save to the repository. Pushing these changes updates the repository on GitHub, making them available to collaborators.

Steps:

1. Make changes to the files in your local repository.
2. Open GitHub Desktop, where your changed files will be listed.
3. Write a clear and descriptive commit message summarizing your changes (e.g., "Fixed bug in data processing script").
4. Click `Commit to main` to save the snapshot locally.
5. Push your changes to GitHub by clicking the `Push origin` button.

Commits and pushes are essential for maintaining a history of changes and sharing updates with collaborators. They ensure transparency and accountability in project development.
Note that a commit is a local operation, i.e. it's tracked locally, but not shared on GitHub or with any of your collaborators until you push the changes.

```r
# Placeholder for screenshot of commit and push process
```

---

### Creating a Pull Request

A pull request is a way to propose changes to the original repository. It allows the repository owner to review your modifications before merging them.
It is the method for moving the changes you made in your fork back to the original repository.

Steps:

1. Open your forked repository on GitHub.
2. Click the `Pull requests` tab and then `New pull request`.
3. Select the original repository as the base and your forked repository as the compare.
4. Add a title and description for your pull request explaining the changes you’ve made.
5. Click `Create pull request`.

Pull requests are valuable for collaborative projects, as they allow for feedback and discussion before changes are finalized.

```r
# Placeholder for screenshot of pull request creation
```

---

# Part 2: Introduction to Markdown Documents with Quarto

## What is Markdown?

Markdown is a lightweight markup language used to create formatted text with plain text editors. It’s widely used for documentation because of its simplicity and readability. Quarto extends Markdown with additional features to create professional documents, including reports, presentations, and technical analyses.

If you are familiar with LaTeX, Markdown is a simpler alternative for creating documents. It allows you to focus on content rather than formatting, making it ideal for writing reports, notes, and documentation.

---

## Creating a Quarto Document

Quarto documents are ideal for reproducible research and professional reporting. To create a Quarto document:

1. Install Quarto from [https://quarto.org](https://quarto.org).
2. Open RStudio or another text editor.
3. Create a new Quarto document via `File > New File > Quarto Document`.

A new `.qmd` file will be generated, containing the following structure:

```markdown
---
title: "Document Title"
author: "Author Name"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction
This is a sample Quarto document.

```{r}
# Example of an R code chunk
summary(cars)
```
```

This structure includes metadata (called the YAML header), plain text content, and code chunks for embedding analyses.

---

## Document Structure

### YAML Header
The YAML header defines metadata for the document, such as the title, author, and output format. It’s placed at the top of the file and surrounded by triple dashes (`---`).

### Plain Text and Formatting

Use plain text for content and Markdown syntax for formatting:

- **Headings**: Use `#`, `##`, and `###` for different heading levels. Headings help structure your document.
- **Lists**: Create lists to organize information clearly. Use `-` for unordered lists or numbers for ordered lists. You can create nested lists by indenting (4 spaces per level).
- **Text Formatting**: Use `*italic*` or `**bold**` to emphasize text.

### Code Chunks

Code chunks allow you to embed and execute code within the document. They are enclosed by triple backticks with a language specifier:

```markdown
```{r}
# Example of an R code chunk
plot(mpg ~ wt, data = mtcars)
```
```

Code chunks are essential for integrating analysis directly into the document, ensuring reproducibility.

### Math Notation

Use LaTeX syntax for mathematical expressions. For example:

- Inline math: `$E = mc^2$`
- Display math: `$$\int_0^1 x^2 \, dx = \frac{1}{3}$$`

Math notation is especially useful in technical and scientific reports.

### Fenced Divs

Fenced divs enable advanced formatting and layouts. For example:

```markdown
:::: {.callout-note}
## Note
This is a callout box.
::::
```

---

## Rendering the Document

To render a Quarto document:

1. Save the `.qmd` file.
2. Click the `Render` button in RStudio.

Rendering produces the final output, such as HTML or PDF, making your analysis and documentation accessible.

